---
title: "Basketball Plots"
author: "Samantha Rodriguez"
date: "1/26/2022"
output: pdf_document
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(devtools)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(STA486Project1sr2354)
library(knitr)
```

# **1 Introduction**
Basketball players have a lot of statistics about their career. Over the years
the game of basketball has changed as well as the playing style. Looking deeper
into players position and performance over the years we can see that change.

# **2 Data & Organization**
To analyze basketball player data, I collected the information from a Kaggle
data set[^1]. The data set was organized in a SQLite format. SQLite holds 
different tables of information separatley. There is no manual way of extracting
the data into a usable R format. In order change the SQLite file into separate 
CSV files of the tables it has, I needed to download an IDE of SQLite called
SQLiteStudio[^2]. The file had sixteen separate tables. I extracted nine CSV
files.

After extracting, I can now import the files to clean and organize the different
data sets. The data sets included information on the players, the teams, the
draft, the draft combine, and the team and player salary. Looking into each 
data set closer, I decided to analyze three of the tables. These will be looking
at the players, draft, and draft combine. After factoring some columns and 
changing the names of the columns to match, they were merged together. Begining
with the `draft` table, I dropped all the `NAs` from the rows. I will be using
this table as the main data set to combine the other two tables. The data set
being 7890 columns and 16 columns. Moving to the next table `draft_combine`
I dropped columns that weren't necessary or had over a 90% `NA` percentage and
then again dropping the `NAs` from all the columns. Then merging the two tables.
The last table used, `player_attributes`, I only selected four of the twenty 
columns and merged it with the final data set.

Leaving the final table is a 217 rows and 30 columns. Some of the columns basis 
are to be connected with the other tables, such as the `idTeam` and `idPlayer`. 
These columns will not be used in the further analysis but it being kept in case 
further changes needed to be made and the tables need a similar column. Other
columns such as the name of the player and the team the played for are not
included in the analysis but are kept for reference. **Table 1** includes all
the columns that were kept. The next section, will go further into the analysis 
being made of the data that has remained.

# **3 Models & Analysis**
Height is a large factor in basketball players, even in player positions that
don't need to be tall. ***Figure 1*** shows the height of the basketball players 
that went through the combine from 2005-2020. The players are separated by what
position they play. There are some players who play up to two positions and 
they are put into their own category. 

[^1]: https://www.kaggle.com/wyattowalsh/basketball
[^2]: https://sqlitestudio.pl/

\newpage

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#knitr::kable(str(player_info))
```


\newpage

```{r}
levels(player_info$yearDraft) <- c("1987", "2004", "2005", "2006", "2007", "2008",
                                   "2009", "2010", "2011", "2012", "2013", "2015",
                                   "2017", "2018", "2019", "2020")
player_info$yearDraft <- as.Date(player_info$yearDraft, "%Y")
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
levels(player_info$slugPosition) <- c("C", "C-PF", "PF", "C-PF", "PF-SF", "PG",
                                      "PG-SG", "SF", "PF-SF", "SF-SG", "SG", 
                                      "PG-SG", "SF-SG")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(player_info, aes(x=slugPosition, y=heightWOShoesInches)) + geom_boxplot() +
  ggtitle("Heights of Different Basketball Positions") +
  labs(x = "Position", y = "Height (inches)", 
       caption = "Figure 1. Height of basketball players of different positions over the years of 1947 and 2020. If a player played multiple 
       positions they are in a group together and not put in each seperate group.") +
  theme(plot.caption = element_text(hjust=0, face="italic"))
```

\newpage

The box plots show that there is clear pattern based on the player height and 
what position they play. With center having the tallest height because they need
to be the biggest players on the court since they're the closest to the basket.
Also, point guards, being the shortest player, because they are doing most of
the dribbling during the plays.

Now, that we have looked at the height of the players, lets look at the weight.
***Figure 2*** displays box plots of the weight of the players based on their 
position. Similar to ***Figure 1***, players who play multiple positions are
put in their own category. Based off these two graphs, we can see that height
and weight are important for basketball players. The farther the player is 
positioned from the basket the shorter and less they weigh. 



The strategy of basketball is each player bring their own set of skills based
on their athletic ability and combine them as a team to beat the other team.
The three main skills needed for basketball is rebounds, assist, and point 
scoring. The different positions favor each skill differently. 

### **Rebounds**

Centers and power
forwards need to be the players rebounding the most on the court. Whether that
be offensively or defensively. As mentioned before, height is important to these
two positions as well. ***Figure 3*** compares the height with their career
average rebounds over the years the players were drafter. We're also looking at 
the height of the players. They are separated by whether they are above or below
the median height of the positions selected. The graph also includes the players 
that play both center and power forward positions.

Ignoring when the height of the player, there is a decreasing
trend on the average rebounds of center and power forwards. There was a steady 
trend on the average number of rebounds in the early 2000's but moving into the
2010's the trend is decreasing. It seems center and power forwards don't
rebound as much as before. Now looking at the height, with red circle being
above median height and blue triangle being below. Visually, there is no 
immediate trend. The are outliers for both sides, but they're seems to be more
red circles averaging a higher rebounds per game. This does make sense as
taller players have a better chance of rebounding.

### **Points**

The next position we'll being looking at his shooting guards. These players are 
supposed to score the most of the team. This is the largest group of players
since they are in the middle of the field. ***Figure 4*** includes shooting
guards, point guards, and shooting forwards. Since, these are the scorers of the
team we will be comparing the average points per game in the players career and
what year they were drafted, as well as which round they were drafted in. There
are two rounds in the NBA draft, the better players are drafted from the first 
round. Because of this we would expect them to perform better. 

Similar to ***Figure 3***, ***Figure 4*** shows a negative trend overall will a
couple bumps in 2008 specifically. In the 2000's shooting guards had a high 
average of points per game. Now in the 2010's, the average has decreased. 
Looking at which round the players were drafted there is a clear patter of the
first round picks averaging more than the second round picks, as expected.
All the shooting guards in the 2010's average what the second round picks did 
in the early 2000's. In the 2010's there is also less first round picks that
are shooting guards than second round picks.

### **Assists**

The last skill is assists, which is given to the point guard position. For
***Figure 5*** we will be including the players who played point guard and 
shooting guard. The graph compares the players assists average of their career
and the year they were drafted as well as what round they were drafted in. 

\newpage

```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(player_info, aes(x=slugPosition, y=weightLBS)) + geom_boxplot() +
  labs(title = "Weight of players in different poistions", 
       x = "Position", y = "Weight (lbs)", 
       caption = "Figure 2. Weight of players in different poistions in the NBA from the year of 1947 to 2020. If a player played in two 
       different poisitons they are in a seperate group and not put in two seperate groups at the same time.") +
  theme(plot.caption = element_text(hjust=0, face="italic"))
```

\newpage

```{r, echo=FALSE, warning=FALSE, message=FALSE}
player_info$numberRound <- factor(player_info$numberRound, levels = c("1", "2"))
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
center.height <- player_info[player_info$slugPosition %in% c("C", "C-PF", "PF-C"),]
med <- median(center.height$rebAvg)
reb01 <- ifelse(center.height$rebAvg > med, 1, 0)
reb01 <- factor(reb01)
center.height <- cbind(center.height, reb01)
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(center.height, aes(x=yearDraft, y=heightWOShoesInches)) + 
  geom_point(aes(color=reb01, shape=reb01), size=2) +
  ggtitle("Heights of Center Poistion Over the Years") +
  labs(caption = "Figure 3. Comparing the center positions height as well as if they are above
       or below the median rebounds.", 
       x="Draft Year", y="Height (inches)") + geom_smooth(se=F) + 
  scale_x_date(date_labels = "%Y", date_breaks = "1 year")
```

\newpage

```{r, echo=FALSE, warning=FALSE, message=FALSE}
shooting.guard <- player_info[player_info$slugPosition %in% c("PG-SG", "SF-SG", "SG", "SG-PG", "SG-SF"),]
ggplot(shooting.guard, aes(x=yearDraft, y=ptsAvg)) + 
  geom_point(aes(color=numberRound, shape=numberRound)) +
  ggtitle("Average Points for Shooting Gaurds") +
  labs(caption = "Figure 4. Comparing the average points of shooting guards also seeing in what
       round they player was drafted.", 
       x="Draft Year", 
       y= "Career Average Points") + geom_smooth(se=F)+ 
  scale_x_date(date_labels = "%Y", date_breaks = "1 year")
```

\newpage

```{r, echo=FALSE, warning=FALSE, message=FALSE}
point.guard <- player_info[player_info$slugPosition %in% c("PG-SG", "PG", "SG-PG"),]
ggplot(point.guard, aes(x=yearDraft, y=astAvg)) + 
  geom_point(aes(color=numberRound, shape=numberRound)) +
  ggtitle("Average Assists for Point Gaurds") +
  labs(caption = "Figure 5. Comparing the average assists of point guards also seeing in what
       round they player was drafted.", 
       x="Draft Year", 
       y= "Career Average Assists") + geom_smooth(se=F)+ 
  scale_x_date(date_labels = "%Y", date_breaks = "1 year")
```

\newpage

Similar to the previous two graphs, there is a negative trend
overall but a bump on 2008-2009. As before, round 1 players averaged more assists
than players drafted in round 2. Another similarity, there is not many point
guards and shooting guards drafted from round 1 in the 2010's as compared to the
2000's. 

The last figure, ***Figure 6*** is the conclusion of these models. The graph is
a stacked bar chart of the different positions and the draft year. Specifically, 
the number of players drafted for each year. This data set does not include all
years which is why there are some blanks in the x-axis. For the years, that are
included, the positions that require more height are drafted more than the other 
positions. In the 2010's the number of positions that have shorter players are
drafted in a very small amount or almost none. As compared to the taller positions
who stay relevant for the full timeline. We will discuss this further in the 
next section.

# **Discussion of Major Results**
The game of basketball changes over the years. From the models listed above it 
is clear that is true. There wasn't as much athletic ability in tall player,
thus the need for shorter players who were faster and more agile. Looking at
the figures, point guards and shooting guards are doing less and less for the 
game. There was a peak of both players in 2008 and 2009. There seems to be less
of a demand for those positions. 2008 and 2009 includes the best players of 
those positions ever since. A couple of those players are Derrick Rose and 
Russell Westbrook, thought to be the best point guards of that time[^3]. The
game is changing and taller players can be faster and more agile. This 
correlates to the change in athletic training that has been made. Different
training regimines has created players that have changed how the game is 
played and how teams carry out their strategy.

There has always been a difference in the performance of a round 1 and round 2
player. As one is supposed to be better than the other. In the 2000's the graph
shows that the statement is true. Round 1 players performed better in their
respective positions. Shooting guards scored more points and point guards had 
more assists. Moving into the 2010's that trend is not the same. Those positions
were drafted more in round 2 than in round 1. Backing up the previous statement,
that those positions aren't as important to the strategy anymore. Players have 
become more well rounded in skills and a specified skill set is not seen as
helpful.

This can be seen in ***Figure 6***, the positions that are mixed are drafted more
than specialized positions. A player that can play a shooting forward and a 
shooting guard is better for the team than a shooting guard. Teams are looking
for players that can do the three main skills needed for the game not just one.

While, not every year was included in the analysis, there was enough of a trend
that the years wouldn't have affected the analysis of the graphs in the end. 


[^3]: http://www.espn.com/nba/draft2008/index

\newpage


```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(player_info, aes(x=yearDraft, y=frequency(slugPosition), fill=slugPosition)) +
  geom_bar(stat = "identity") +
  labs(title = "Frequency of Player Positions Drafted Over the Years",
       x = "Year", y="Frequency",
       caption = "Figure 6. The frequency of how each poistion changes importance over the years.")+ 
  scale_x_date(date_labels = "%Y", date_breaks = "1 year")
```

\newpage

# **Conclusion**
The game of basketball has a strategy to win just like any game. The strategy
may change and the players and teams have to adjust accordingly. The results
showed this change. Well rounded players have a higher chance of being drafted
than single specialized players. With the advancement in sports medicine, 
players can become more well rounded.


# **References**
1. https://www.kaggle.com/wyattowalsh/basketball
2. https://sqlitestudio.pl/
3. http://www.espn.com/nba/draft2008/index
















